### 2025-05-03 Chapters/Chapter2.tex
komið lengra

Data from these automatic weather stations (AWS) are stored in hourly files, which aggregate the original 10-minute files; measurement errors—unrealistic spikes known as “nails”—have been removed in most cases. Each record contains the following information: date and time; station number (convertible to coordinates using another dataset of Icelandic meteorological stations); average wind speed (\(f\)); wind gust (\(f_g\)); standard deviation of the wind gust; wind direction (\(d\)); and standard deviation of the wind direction.

@@

\section{Elevation data}
A GeoTIFF file containing Iceland’s elevation on a 20~m by 20~m grid was provided by the Icelandic Meteorological Office (IMO). The entire country is covered by this file, and its size is approximately 685\,MB.

@@

Elevation data are handled by defining an upwind sector: a range of angles relative to the wind direction \(d\) and radial distances from each station. Points within this sector are generated as shown in Code Listing \ref{code:sectorElevation}.

@@

\begin{table}[h]
    \centering
    \caption[Example of combined data structure]{Example of the combined data structure used for modeling. Data include derived variables Richardson number (\(Ri\)) and squared Brunt–Väisälä frequency (\(N^2\)), station elevation, wind direction and relative wind direction (twd), along with wind speed, pressure, and temperature at various height levels, and elevation points around each station relative to station elevation.}
    \label{table:trainDataExample}
    \resizebox{\textwidth}{!}{
    \begin{tabular}{ccccccccccc}
        \toprule
        Ri & \(N^2\) & \shortstack{station\\elevation} & twd & ws\(_{15}\) & wd\(_{15}\)
        & \(t_{15}\) & \(p_{15}\) & \(\text{elevation}_0\) & elevation\(_1\) & \dots\\
        \midrule 
        -1.18 &  26700 & 100 & 1.5 & 10 & 5 & 0 & 100 & 2 & 4 & \dots\\
        \dots\\
        \bottomrule
    \end{tabular}
    }
\end{table}

In Table \ref{table:trainDataExample}, the first two columns describe atmospheric stability: the Richardson number (\(Ri\)) and squared Brunt–Väisälä frequency (\(N^2\)), defined in Equations (\ref{eqn:Ri}) and (\ref{eqn:N}) \cite{richardson_number_skybrary,brunt_vaisala_freq_eumtrain,mean_gust_HA_HO}. Here, \(g\) is gravitational acceleration; \(dT\) the temperature difference between levels; \(dz\) the height difference; \(T_{\mathrm{ave}}\) the average temperature; and \(dU\) the wind-speed difference. Lower \(Ri\) indicates greater turbulence, with typical values between 0.1 and 10, and values below 1 signifying significant turbulence \cite{richardson_number_skybrary}. Negative \(N^2\) denotes instability, as an air parcel will accelerate away from its original position \cite{brunt_vaisala_freq_eumtrain}.

Since \(Ri\) and \(N^2\) are derived from reanalysis variables, they add limited new information compared to the raw data, but they can reduce input dimensionality and improve model explainability via Shapley-value analysis \cite{shapley_information}. Shapley values assess feature importance by evaluating all possible feature subsets—an operation of computational complexity \(2^n\). In practice, approximations are used but can still be costly for large models. Because \(dU\) appears squared in the denominator of \(Ri\), if the wind-speed difference between levels is very small, \(Ri\) can become arbitrarily large, potentially distorting predictions.

@@

\begin{figure}[ht]
  \centering
  \includegraphics[width=0.8\textwidth]{Figures/station_heights.png}
  \caption{Distribution of weather station elevations above sea level. One outlier near 2000~m was excluded due to limited and inconsistent data.}
  \label{fig:station_heights}
\end{figure}

\begin{figure}[ht]
  \centering
  \includegraphics[width=0.8\textwidth]{Figures/gust_factor_2025.png}
  \caption{Histogram of gust factors. By definition, gust factor \(\ge1\); most values lie between 1.2 and 2. The gust factor declines as wind speed increases.}
  \label{fig:gust_factors}
\end{figure}

>>>>

Data from these automatic weather stations (AWS) are stored in hourly files, which aggregate the original 10-minute files; measurement errors—unrealistic spikes known as “nails”—have been removed in most cases. Each record contains the following information: date and time; station number (convertible to coordinates using another dataset of Icelandic meteorological stations); average wind speed ($f$); wind gust ($f_g$); standard deviation of the wind gust; wind direction ($d$); and standard deviation of the wind direction.

@@

\section{Elevation data}
A GeoTIFF file containing a digital elevation model (DEM) for Iceland on a 20~m by 20~m grid was provided by the Icelandic Meteorological Office (IMO). The entire country is covered by this file, and its size is approximately 685\,MB.

@@

The DEM elevation data are handled by defining an upwind sector: a range of angles relative to the wind direction $d$ and radial distances from each station. Points within this sector are generated as shown in Code Listing \ref{code:sectorElevation}.

@@

\begin{table}[h]
    \centering
    \caption[Example of combined data structure]{Example of the combined data structure of features used for modeling. Data include derived variables Richardson number (Ri) and squared Brunt–Väisälä frequency ($N^2$), station altitude (meters above sea level), transformed wind direction (twd), wind speed (ws$_{15}$), wind direction (wd$_{15}$), temperature ($t_{15}$), pressure ($p_{15}$) (CARRA values at 15 m height), and elevation points (from DEM) in a sector pointing upwind.}
    \label{table:trainDataExample}
    \resizebox{\textwidth}{!}{
    \begin{tabular}{ccccccccccc}
        \toprule
        Ri & $N^2$ & \shortstack{station\\altitude} & twd & ws$_{15}$ & wd$_{15}$
        & $t_{15}$ & $p_{15}$ & $\text{elevation}_0$ & elevation$_1$ & \dots\\
        \midrule 
        -1.18 &  26700 & 100 & 1.5 & 10 & 5 & 0 & 100 & 2 & 4 & \dots\\
        \dots\\
        \bottomrule
    \end{tabular}
    }
\end{table}

The transformed wind direction is a proxy showing whether the wind is coming from land or sea. It is computed as the angle between the wind direction and a vector from the center of Iceland to the station.

The Richardson number and the Brunt–Väisälä frequency describe atmospheric stability, see equations (\ref{eqn:Ri}) and (\ref{eqn:N})  \cite{richardson_number_skybrary,brunt_vaisala_freq_eumtrain,mean_gust_HA_HO}. These values are calculated using reanalysis data at two different height levels. Thus Ri refers to the Richardson number calculated between height levels 15m and 500m. Exactly the same notation is used with the Brunt–Väisälä frequency, except the square is used.
%
\begin{equation}
  \label{eqn:Ri}
  Ri = \frac{g \cdot dT \cdot dz}{T_{\textrm ave} \cdot dU^2} \unit{[]}
\end{equation}
%
\begin{equation}
  \label{eqn:N}
  N = \sqrt{\frac{g \cdot dT }{T_{\textrm ave} \cdot dz}} \unit{[Hz]}
\end{equation}
%
Here, $g$ is gravitational acceleration; $dT$ the temperature difference between levels; $dz$ the height difference; $T_{\mathrm{ave}}$ the average temperature; and $dU$ the wind-speed difference. Lower Ri indicates greater turbulence, with typical values between 0.1 and 10, and values below 1 signifying significant turbulence \cite{richardson_number_skybrary}. Negative $N^2$ denotes instability, as an air parcel will accelerate away from its original position \cite{brunt_vaisala_freq_eumtrain}.

Since Ri and $N^2$ are derived from reanalysis variables, they add limited new information compared to the raw data, but they can reduce input dimensionality and improve model explainability via Shapley-value analysis \cite{shapley_information}. Shapley values assess feature importance by evaluating all possible feature subsets—an operation of computational complexity $2^n$. In practice, approximations are used but can still be costly for large models. Because $dU$ appears squared in the denominator of Ri, if the wind-speed difference between levels is very small, Ri can become arbitrarily large, potentially distorting predictions.

@@

\begin{figure}[ht]
  \centering
  \includegraphics[width=0.8\textwidth]{Figures/station_heights.png}
  \caption{Distribution of weather station altitudes above sea level. One outlier near 2000~m was excluded due to limited and inconsistent data.}
  \label{fig:station_heights}
\end{figure}

\begin{figure}[ht]
  \centering
  \includegraphics[width=0.8\textwidth]{Figures/gust_factor_2025.png}
  \caption{Histogram of gust factors. By definition, gust factor $\ge1$; most values lie between 1.2 and 2. The gust factor declines as wind speed increases.}
  \label{fig:gust_factors}
\end{figure}

### 2025-05-03 Chapters/Chapter4.tex
Update Chapter4.tex
Skipti kafla 4 í þrjá undirkafla.
komið lengra

\chapter{Results}
\label{Chapter4}
\section{Results}
A baseline model was constructed. This model looked at the gust factor for some training data and took the average of this and predicted this average everytime. Several baseline guesses were created based on a lower limit assigned to the average wind speed limit (AWSL). As the average wind speed increases, then the variability in gust as a percentage decreases\cite{mean_gust_HA_HO}. This means looking at a subset of the data where the AWSL is higher, a better result can be expected. The results show this. Several different AWSL were used for the baseline model, as can be seen in table \ref{table:results}. This sets a goal. A model that does not significantly improve on this baseline suggests either failure to capture essential patterns in the data or that the data itself may lack the necessary information for substantial improvements upon the baseline. Using the previously described neural network architecture setups for each AWSL, with and without landscape elevation information, MAPE was determined. The results can be seen in Table \ref{table:results}.

@@

@@

@@

\begin{table}[h]
    \caption[Model results for different sets of parameters.]{Showing the results of different models. Models defined by different sets of parameters, with increasing complexity. This table shows the feature importance as the loss decreases as more variables are added. Adding all height levels decreases the loss but very little.}
    \label{table:setsOfParams}
    \centering
    \begin{tabular}{lc}
        \toprule
        \textbf{Model parameters} & \textbf{MAPE}\\
        \midrule
        Baseline constant & 23.9\%\\
        $ws_{15}$ & 17.2\%\\
        $[ws_{15}, t_{15}, p_{15}, wd_{15}]$ & 16.7\% \\
        $[ws_{15}, t_{15}, p_{15}, wd_{15}, ASL, twd]$ & 13.8\% \\
        $[ws_{15}, t_{15}, p_{15}, wd_{15}, ASL, twd, N, Ri]$ & 13.7\%\\
        $[ws_{15}, t_{15}, p_{15}, wd_{15}, ASL, twd, N, Ri]$ + DEM & 12.0\%\\
        $[ws_{15,250,500}, t_{15,250,500}, p_{15,250,500}, wd_{15,250,500}, ASL, twd, N, Ri]$ & 12.0\% \\
        \bottomrule
    \end{tabular}
\end{table}

>>>>

\chapter{Results}
\label{Chapter4}
\section{Baseline models}
The simplest model for the gust factor is to guess the average gust factor for all available observations, i.e. use no variables from the training data. A slightly better model is obtained by using the wind speed as a training feature, either through regression or NN training. The MAPEs for the constructed base models are shown in Table \ref{table:baseline_models}. The ws$_{15}$ model was obtained with NN training but regression produces an almost identical result.

\begin{table}[h]
  \caption[MAPE for baseline models]{MAPE for several lower limits on the reanalysis 15 m wind speed ($ws_{15}$). Note that the ws$_{15}$ model was only evaluated for the 10 m/s lower bound.}
    \label{table:baseline_models}
    \centering
    \begin{tabular}{lcc}
        \toprule
        Wind speed & \multicolumn{2}{c}{MAPE}\\ 
        limit $[m/s]$ & mean model & ws$_{15}$ model\\
        \midrule
        $\geq 0$ & 39.2\%  & \\
        $\geq 5$ & 28.1\%  & \\
        $\geq 10$ & 23.9\% & 17.2\%\\
        $\geq 15$ & 23.2\% & \\
        $\geq 20$ & 24.7\% & \\
        $\geq 25$ & 27.7\% & \\
        \bottomrule
    \end{tabular}
\end{table}

As the average wind speed increases, then the variability in gust as a percentage decreases\cite{mean_gust_HA_HO}. Therefore lower MAPE woult be expected if the training data is restricted to higher wind speeds. This fact explains why several models were created for different lower bounds.

\section{Modeling for CARRA wind speed above 10 m/s}
As noted in Chapter 1, it is more important to predict accurate gust factor when it is windy than when the weather is calm. Therefore, at first, the data is restricted to ws$_{15} \geq 10$ m/s. The first constructed model incorporates the features temperature, pressure and wind direction at 15 m height. This gives a moderately reduced MAPE of 16.7\% compared with 17.2\% for the baseline ws$_{15}$ model. When the next pair of features, station altitude and transformed wind direction, is included, the MAPE is reduced considerably, down to 13.8\%. The atmospheric stability indicators, Ri and $N^2$, add little to the model performance, contrary to what might have been expected. When the atmosphere is unstable, wind gusts from higher up might be expected. However, another considerable perfomance increase is observed when digital elevation model (DEM) information is included, down to a MAPE of 12.0\%. This represents over 30\% reduction in the MAPE from the ws$_{15}$ baseline model.

\begin{table}[h]
    \caption[Model results for different sets of parameters.]{Showing the results of different models. Models defined by different sets of parameters, with increasing complexity. This table shows the feature importance as the loss decreases as more variables are added. Adding all height levels decreases the loss but very little.}
    \label{table:setsOfParams}
    \centering
    \begin{tabular}{lc}
        \toprule
        \textbf{Model variables} & \textbf{MAPE}\\
        \midrule
        Baseline constant & 23.9\%\\
        $ws_{15}$ & 17.2\%\\
        $[ws_{15}, t_{15}, p_{15}, wd_{15}]$ & 16.7\% \\
        $[ws_{15}, t_{15}, p_{15}, wd_{15}, \text{altitude}, twd]$ & 13.8\% \\
        $[ws_{15}, t_{15}, p_{15}, wd_{15}, \text{altitude}, twd, N^2, Ri]$ & 13.7\%\\
        $[ws_{15}, t_{15}, p_{15}, wd_{15}, \text{altitude}, twd, N^2, Ri]$ + DEM & 12.0\%\\
        $[ws_{15,250,500}, t_{15,250,500}, p_{15,250,500}, wd_{15,250,500}, \text{altitude}, twd, N^2, Ri]$ & 12.0\% \\
        \bottomrule
    \end{tabular}
\end{table}

The results show this. Several different AWSL were used for the baseline model, as can be seen in table \ref{table:results}. This sets a goal. A model that does not significantly improve on this baseline suggests either failure to capture essential patterns in the data or that the data itself may lack the necessary information for substantial improvements upon the baseline. Using the previously described neural network architecture setups for each AWSL, with and without landscape elevation information, MAPE was determined. The results can be seen in Table \ref{table:results}.

@@

\section{Results for specific locations}

@@

\section{Comparison for models with variable parameters}

@@
